---
title: "Hazards"
date: "Generated: `r Sys.Date()`"
runtime: shiny
output:
    html_document:
        toc: true
        toc_depth: 3
        theme: spacelab
        highlight: zenburn
        number_sections: false
---

```{r global_opts, include=FALSE}

knitr::opts_chunk$set(fig.path="Figs/", echo=FALSE, warning=FALSE,
                      message=FALSE, dev='svg')

```

-------

# Info
* [code](https://github.com/nverno/hazards)

Simple logistic models to predict death.

-------

# Moose Data
`DIED` variable is all trees that were alive in first census period and __not reported__ as alive in next period.  `rDIED` variable is all trees that were actually __reported__ to have status "DEAD" or "PD".

```{r summary}
library(ggplot2)
library(GGally)
library(dplyr)
library(grid)
library(reshape2)
source("setup.R", chdir=T)

cols <- c("#A60628", "#348ABD")
specs <- levels(pp$SPEC)[!(levels(pp$SPEC) %in% c("", "UNID"))]
diedCols <- c("DIED", "rDIED")

inputPanel(
    checkboxGroupInput("spec", "Species:", choices=specs, selected=c("ABBA"), inline=TRUE),
    checkboxGroupInput("times", "Time Periods:", choices=c(86, 87, 98), selected=c(98), 
                       inline=TRUE),
    checkboxGroupInput("groups", "Split Table:", choices=c("SPEC", "YEAR", "ASPCL", "ELEVCL", "PPLOT"),
                       selected=c("SPEC", "YEAR"), inline=TRUE),
    checkboxInput("treesOnly", "Trees >=5cm Only")
)

getData <- reactive({
    inds <- pp$SPEC %in% input$spec & pp$YEAR %in% input$times
    if (input$treesOnly)
        inds <- inds & !is.na(pp$DBH) & pp$DBH >= 5
    droplevels(pp[inds, ])
})

actionButton("deathSummary", "Show Summary of Deaths")

renderTable({
    if (input$deathSummary %% 2) {
        dname <- ifelse (input$treesOnly, "Total(>=5cm)", "Total")
        dat <- getData()
        props <- dat %>% group_by_(.dots=input$groups) %>%
          dplyr::summarize(
              "Died" = as.integer(sum(DIED, na.rm=T)),
              "Total" = n(),
              "Died(%)" = mean(DIED, na.rm=T),
              "rDied(%)" = mean(rDIED, na.rm=T))
        props
    }
})

summPlot <- renderPlot({
    dat <- getData()
    dcols <- c("dp", "rdp")[c("DIED", "rDIED") %in% input$summDvar]
    svars <- c("SPEC", input$summGroup)
    if (input$summSplit) svars <- unique(c(svars, input$summFacet))
    props <- dat %>% 
      group_by_(.dots=as.list(svars)) %>%
      dplyr::summarize(dp = mean(DIED, na.rm=T),
                       rdp = mean(rDIED, na.rm=T))
    props <- melt(props, id.vars = svars)
    props <- props[props$variable %in% dcols, ]
    p <- ggplot(props, aes_string(input$summGroup, "value", group="SPEC", shape="variable", color="SPEC")) + 
      geom_point() + theme_bw() +
      ylab("Proportion Died") +
      scale_shape_discrete(name="Died Variable", labels=c("DIED", "rDIED"))
    if (input$summPath)
        p <- p + geom_path(aes(group=interaction(SPEC, variable)), arrow=arrow())
    if (input$summSplit) 
        p <- p + facet_grid(as.formula(paste("~ ", input$summFacet)))
    p
})

shinyUI(fluidPage(
    titlePanel("Deaths Summary Figures"),
    sidebarLayout( 
        sidebarPanel(
            checkboxGroupInput("summDvar", "Died variable:", choices=c("DIED", "rDIED"), 
                               selected="DIED", inline=T),
            radioButtons("summGroup", "Groups:", 
                         choices=c("PPLOT","ASPCL","ELEVCL","YEAR"), 
                         selected=c("YEAR"), inline=T),
            checkboxInput("summPath", "Add Path"),
            checkboxInput("summSplit", "Split Graphs"),
            conditionalPanel(
                condition = "input.summSplit == true",
                radioButtons("summFacet", "Facets:", choices=c("YEAR", "ELEVCL", "ASPCL", "PPLOT"), 
                             selected=NULL, inline=TRUE)
            )
        ),
        mainPanel(
            summPlot
        )
    )
))

```

-------

# Plot-Variable Correlations

```{r plot_corrs}

corrHist <- renderPlot({
    dat <- getData()
    hist(dat[, input$corrVar], main=paste("Histogram of", input$corrVar),
         xlab=input$corrVar)
})

corrGroups <- renderPlot({
    dat <- getData()
    dat <- dat[dat$PPLOT %in% input$corrPlots, ]
    p <- ggplot(dat, aes_string(input$corrVar)) + geom_histogram()
    if (input$corrRotate)
        p <- p + facet_grid(~ PPLOT)
    else
        p <- p + facet_grid(PPLOT ~ .)
    p
})

corrPairs <- renderPlot({
    dat <- getData()
    dat <- dat[dat$PPLOT == input$pairPlot, input$pairVars]
    ggpairs(dat, upper=list(continuous="density"),
            lower = list(combo="facetdensity")) + theme_bw()
})

shinyUI(fluidPage(
    titlePanel("Correlations between PPLOT and variables"),
    sidebarLayout(
        sidebarPanel(
            conditionalPanel(
                condition = "input.corrTabs == 'Histogram' || input.corrTabs == 'Split'",
                radioButtons("corrVar", "Variable:", choices=names(pp), 
                             selected="BA", inline=TRUE),
                conditionalPanel(
                    condition = "input.corrTabs == 'Split'",
                    checkboxGroupInput("corrPlots", "Plots:", choices=levels(pp$PPLOT), 
                                       selected=c(4,5,6), inline=TRUE),
                    checkboxInput("corrRotate", "Rotate Split")
                )
            ),
            conditionalPanel(
                condition = "input.corrTabs == 'Pairwise'",
                radioButtons("pairPlot", "Plot:", choices=levels(pp$PPLOT), 
                             selected=4, inline=TRUE),
                checkboxGroupInput("pairVars", "Variables:", choices=names(pp), 
                             selected=c("HT","BA"), inline=TRUE)
            )
        ),
        mainPanel(
            tabsetPanel(
                id = "corrTabs",
                tabPanel("Histogram", corrHist),
                tabPanel("Split", corrGroups),
                tabPanel("Pairwise", corrPairs)
            )
        )
    )
))

```


-------

# Simulating Data

## Right-censored

* Simulate some lifetimes (exponential, gamma)
* Simulate some termination (censoring) times (uniform)

Simulate some right-censored data.  Lifetimes for two groups from __exponential__ distributions with different rate paramters.  Censoring times are generated from a __uniform__ distribution, with a lower bound set to the mean of the first groups distribution (1/Rate1):
```{r simulate_data}
library(survival)
data(aml)

inputPanel(
    numericInput("rightSimN", "Number of Samples (per group):", 50, 25, 100),
    numericInput("rightSimRate1", "Rate (Group 1):", 0.2, 0.05, 0.95),
    numericInput("rightSimRate2", "Rate (Group 2):", 0.1, 0.05, 0.95),
    checkboxInput("rightSimSumm", "Summary Table")
)

rightSimData <- reactive({
    lifetimes <- c(rexp(ceiling(input$rightSimN), rate=input$rightSimRate1),
                    rexp(ceiling(input$rightSimN), rate=input$rightSimRate2))
    censtimes <- rep(1/input$rightSimRate1*(1 + runif(input$rightSimN)), 2)
    ztimes <- pmin(lifetimes, censtimes)
    status <- as.numeric(censtimes > lifetimes)
    group <- factor(rep(c(1,2), each=input$rightSimN))
    data.frame(lifetimes=lifetimes, censtimes=censtimes, ztimes=ztimes, 
             group=group, status=status)
})

fluidRow(
    actionButton("rightSimTable", "Show Some Simulation Data"),
    actionButton("runRightSim", "Run Right-censored Simulation",
                 style="background:red; color:black; font-weight:bold")
)

renderTable({
    if (input$runRightSim & input$rightSimTable %% 2){
        dat <- rightSimData()
        if (input$rightSimSumm)
            summary(dat)
        else head(dat)
     }
})

rightSimFit <- reactive({
    if (input$runRightSim) {
        dat <- rightSimData() 
        survfit(Surv(dat$ztimes, dat$status) ~ dat$group,
                conf.int=input$rightCI, conf.type=input$rightCItype)
    }    
})

rightKM <- renderPlot({
    if (input$runRightSim) {
        fit <- isolate(rightSimFit())
        dat <- rightSimData()
        plot(fit, main=paste("Kaplan-Meier with", round(input$rightCI, 2), 
                      "confidence bands"))
        if (input$rightAddConf) {
            for (i in 1:2) {
                lines(fit$time[dat$group==i], fit$upper[dat$group==i], lty=3, type="s")
                lines(fit$time[dat$group==i], fit$lower[dat$group==i], lty=3, type="s")
            }
        }
    }
})

rightKMresultsFull <- renderPrint({
    if (input$runRightSim) {
        fit <- isolate(rightSimFit())
        summary(fit)
    }
})

rightKMresults <- renderPrint({c
    if (input$runRightSim) {
        fit <- isolate(rightSimFit())
        fit
    }
})

rightKMcompare <- renderPrint({
    if (input$runRightSim) {
        dat <- isolate(rightSimData())
        fit <- survdiff(Surv(dat$ztimes, dat$status) ~ dat$group)
        fit
    }
})

shinyUI(fluidPage(
    titlePanel("Right-Censored Simulated Data"),
    sidebarLayout(
        sidebarPanel(
            sliderInput("rightCI", "Conf. level:", min=0.5, max=0.99, value=0.95),
            radioButtons("rightCItype", "Conf. Type:", choices=c("plain","log","log-log"), 
                         selected=NULL, inline=TRUE),
            conditionalPanel(
                condition = "input.rightSimTabs == 'Kaplan-Meier'",
                checkboxInput("rightAddConf", "Show Conf. Bands")
            )
	),
        mainPanel(
            tabsetPanel(
                id = "rightSimTabs",
                tabPanel("Kaplan-Meier", rightKM),
                tabPanel("Results", rightKMresults),
                tabPanel("Full Results", rightKMresultsFull),
                tabPanel("Compare Groups", rightKMcompare)
            )
        )
    )
))
```

```{r test}
library(survival)
lifetimes <- c(rexp(25, rate=0.2), rexp(25, rate=0.1))
censtimes <- rep(5 * (1 + runif(25)), 2)
ztimes <- pmin(lifetimes, censtimes)
status <- as.numeric(censtimes > lifetimes)
group <- rep(c(0, 1), each=25)
dat <- data.frame(lifetimes=lifetimes, censtimes=censtimes,
                  ztimes=ztimes, status=status, group=group)
fit <- survfit(Surv(dat$ztimes, dat$status) ~ dat$group, conf.int=.95)
conf <- data.frame(upper=fit$upper, lower=fit$lower, group=dat$group)

```

-------

# Models

A good model is `DIED ~ BA + HT + CPOS + CANHT`, possibly with `PPLOT`, and it is slightly better with `HTOBS` instead of `HT` (`HT` includes predicted heights, sample size changes when using `HTOBS` instead).

## Logistic
```{r logistic}
## Reactives/Observers
getModelData <- reactive({
    dat <- getData()
    na.omit(dat[, c(input$logDied, input$indVars)])
})

getModel <- reactive({
    dat <- getModelData()
    form <- as.formula(paste(input$logDied, "~", paste(input$indVars, collapse="+")))
    glm(form, data = dat, family=binomial(link=input$link))
})

observe({
    updateRadioButtons(session, "xVar", choices=input$indVars, inline=TRUE)
})

## Output
plotMod <- renderPlot({
    x <- input$xVar
    dat <- getModelData()
    mod <- getModel()
    dat$preds <- fitted(mod)
    
    ggplot(dat, aes_string(x, input$logDied)) + 
      geom_point(position=position_jitter(height=0.05), alpha=0.5, col=cols[1]) +
      geom_point(aes_string(x, "preds"), col=cols[2], alpha=0.5) +
      geom_line(aes_string(x, "preds"), col=cols[2], alpha=0.5) + 
      geom_hline(y=0.5, lty=2) + theme_bw() + 
      ggtitle(paste(input$logDied, "vs", paste(input$indVars, collapse=",")))
})

plotHazard <- renderPlot({
    x <- input$xVar
    dat <- getData()
    mod <- getModel()
    dat$hazards <- log(1 - fitted(mod))
    
    plot(dat[,"hazards"] ~ dat[,x])
    ## ggplot(dat, aes_string(x, "hazards")) +
    ## ggplot(dat, aes_string(x, "DIED")) + 
    ##     geom_point(position=position_jitter(height=0.05), alpha=0.5, col=cols[1]) +
            ## geom_point(aes_string(x, "hazards"), col=cols[2], alpha=0.5) +
            ##     geom_line(aes_string(x, "hazards"), col=cols[2], alpha=0.5) + 
            ##         geom_hline(y=0.5, lty=2) + 
            ##             ggtitle(paste("DIED vs", paste(input$indVars, collapse=",")))
})

shinyUI(fluidPage(
    titlePanel("Fit logistic model"),
    sidebarLayout(
        sidebarPanel(
            radioButtons("logDied", "Died Variable:", choices=c("DIED","rDIED"), selected="DIED", 
                         inline=TRUE),
            helpText("pg = 'Prior Growth'"),
            checkboxGroupInput("indVars", "Predictors", 
                               choices=c("BA", "BV", "ELEV", "HT", "HTOBS", "CPOS", "CANHT", 
                                         "pgBA", "pgBV", "pgHT", "PPLOT"), 
                               selected=c("BA", "HT", "CPOS", "CANHT"),
                               inline=TRUE),
            radioButtons("xVar", "X-axis Variable:",
                         choices=c("BA", "BV", "ELEV", "HT", "CPOS", "CANHT", "HTOBS",
                                   "pgBA", "pgBV", "pgHT"), selected=c("BA"),
                         inline=TRUE),
            radioButtons("link", "Link function:",
                         choices=c("cloglog", "logit", "probit"), inline=TRUE)
        ),
    
        mainPanel(
            tabsetPanel(
                id = "tabs1",
                tabPanel("Probabilities", plotMod),
                tabPanel("Hazards(Not Working)", plotHazard)
            )
        )
    )
))

```

### Logistic model statistics
A good [summary](http://stats.stackexchange.com/questions/108995/interpreting-residual-and-null-deviance-in-glm-r) of how to interpret Null and Residual deviance.  Basically, the Null model has one parameter, while the Residual model is the fitted model.  The residual deviance will be about -2*log-likelihood, and a greater difference between the Null deviance and the Residual deviance is a good sign that the predictors are meaningful, and small deviances mean good fit.  The Residual deviance - Null deviance should be roughly Chi-square (asymptotic) with df = residual df - null df.

```{r model_stats}
actionButton("logResults", "Show Logistic Model Results")
renderPrint({
    if (input$logResults %% 2) {
        mod <- getModel()
        print(summary(mod))
    } 
})

```


## Cox Proportional Hazards Models
